Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-12-04 draft.tex
shortened first two sections, created new paragraphs to save some space

\noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP}, gather as much information as possible about participants to be assessed towards different diseases.
%%
This information is used to determine risk factors for diseases, help people to live healthier or to support the diagnosis of widespread diseases.
%%
So far, epidemiological research is strongly hypothesis-driven.
%%
More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
%%
As a result, the hypothesis will be accepted or rejected.
%%

Cohort studies are an instrument of epidemiological research.
%%
Nowadays, these studies often comprise medical image data, e.g., tomographic image data sets, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
%%
Back pain is one of the most frequent complaints in the Western civilization and gains importance due to the population aging \cite{Hoy2010}.
%%
Epidemiological analysis of back pain aims to determine risk factors.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for back pain, an automatic classification approach for characterization of back pain based on lumbar back spine attributes is still missing.
%%
Furthermore, previous analyses mainly focused on a single variable (e.g. \emph{BMI}) or a small subset of variables.
%%

We present an analysis of the image-derived data of a cohort study including lumbar spine datasets.
%%
We extract possible associations between spine shape and back pain characteristics.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
Then, Interactive Visual Analysis (IVA) highlights mutual dependencies between image-derived data and back pain-related variables.
%%
The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data-driven techniques.
%%
We focus on highlighting new unexpected correlations and trigger \emph{hypotheses generation}, rather than statistically validate complex epidemiological correlations.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
%%
\item The identification of lumbar spine shape properties potentially associated with back pain,
%%
\item The detection of associations between image-based, socio-demographic and medical variables for hypothesis generation,
%%
\item The identification of the most important variables via data mining methods using a novel semi- quantitative overview visualization concept embedded in our IVA-framework.
%%
\end{itemize}
%%
The analysis, as well as the presented IVA tool is available as supplementary material in an interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
%%
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item Epidemiologists compile a list of variables depicting a hypothesis to prove it.
%%
	\item Statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age \cite{Fletcher}.
%%
For example, the effect of nutrition on prostate cancer requires age normalization due to the increased incidence for older subjects.
%%
Age acts as a \emph{risk factor} for prostate cancer.
%%
Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
\subsection{Challenges of Epidemiological Data Analysis}
Epidemiological data originates from a wide range of studies.
%%
The study type depends on the condition of interest.
%%
Most common are case-control studies analyzing one specific disease and its influence on the human body.
%%
We focus on large scale cohort study data.
%%
These studies aim at collecting as much data as possible for each subject.
%%
The collected data can be analyzed regarding many diseases and conditions.
%%

Epidemiological data are heterogenous and incomplete.
%%
For example, data about a disease treatment only affects subjects suffering from this condition.
%%
Therefore, statistical analysis has to take missing data into account.
%%

Epidemiological data acquisition includes a variety of techniques, such as medical examinations, self-reported questionnaires or genetic examinations.
%%
This yields a heterogenous information space.
%%
To compare these data, information reduction techniques are necessary.
%%
For example, continuous data, such as age, is often discretized into quantile bins for visualizing age-dependent disease frequencies.
%%

Modern cohort studies often comprise medical image data.
%%
The analysis of these data is challenging.
%%
Individual diagnosis or manual segmentation of each body structure by radiologists is tedious, costly and comprises little reproducibility.
%%
Segmentation algorithms are not generally available and need to be carefully adapted for each body structure.
%%
Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%%
These numeric values can be compared with non-image variables to retrieve correlations.
%%--------------------------------
\subsection{Lower Back Pain}
%%
Lower back pain is one of the most frequent complaints in the Western civilization \cite{Hoy2010}.
%%
Epidemiologists assume correlations between lumbar back pain and lifestyle factors.
%%
These include \emph{nutrition}, \emph{physical activity} and \emph{body posture combined with physical stress at work}.
%%
The exact causes as well as particularly vulnerable risk groups are unknown.
%%

@@

To achieve this, they have to analyze the lumbar spine shape as well as the mentioned lifestyle factors.

@@

\paragraph{Visual Analysis of Medical Image and Non-Image Data}

@@

% TODO: BLIND
\cite{Klemm2014VIS} propose an extension to the previously described epidemiological workflow using IVA.

@@

We incorporate the IVA approach by concurrently displaying complex relationships between variables via decision trees.

@@

\paragraph{Visual Analysis of Heterogenous Non-Image Data.}

@@

\cite{Dai} define a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
The \emph{Concept Map} is iteratively refined using primarily the underlying geographical data.

@@

\paragraph{Decision Rule-Driven Analysis of Medical Data.}

@@

\cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables, such as age, gender and geographic location about the mortality rate in liver cancer data.
%%
\cite{Sekhavat} propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
%%
The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial and a target value.
%%
The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
%%
\paragraph{SHIP Data Analysis.}
\cite{Klemm2013VMV} analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects.
%%
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier clusters.
%%
The extracted lumbar spine shape was correlated with subject size.
%%
Our data extraction process is based on the detection algorithm and lumbar spine canal extraction presented by \cite{Klemm2013VMV}.
%%
Clustering techniques are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
%%
\\\\
Unique in our work is the combination of data mining techniques with an IVA approach by observing interesting feature relations in the context of image-derived features using multiple decision trees for one visualization.
%%
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in an overview visualization.
%%
\section{\uppercase{The Lumbar Spine Data Set}}
\label{sec:MaterialsAndMethod}
%%
\noindent Our approach allows the simultaneous analysis of many variables.
%%
Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.
%%
The data set comprises 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}) including non-image and image-derived data.
%%
\subsection{Non-Image Data}
%%
The non-image variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors, e.g. sporting activity or nutrition.
%%
The data set comprises 134 variables (and 9 additional image-derived variables described in the next subsection):
\begin{itemize}
	\item 21 metric variables, mostly describing somatometric variables and markers retrieved from blood analysis
	%%
	\item 113 categorical variables divided into
	%%
	\begin{itemize}
		%%
		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. \emph{pancreatitis} or \emph{high blood pressure}
		%%
		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
		%%
	\end{itemize}
%%
\end{itemize}
%%
Many variables comprise missing data.
%%
Some variables are follow-up questions, covering the treatment of a disease.
%%
Others are exclusive to subgroups, such as women-specific variables, e.g. \emph{menstrual status}.
%%
Sparse variables are statistically less resilient and need to be treated with care during the analysis.

\subsection{Image-Derived Data}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.475\textwidth]{figures/centerline}
  \caption{
	(a) left: Tetrahedron-based finite element model (FEM) of the lumbar spine, capturing the L1-L5 vertebrae as well as the lumbar spine canal.
	%%
	(a) right: FEM aligned to an MRI data set.
	%%
	The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
	%%
	(b) Using the Frenet frame, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).
	%%
	The curvature angle ($\alpha$) is calculated using the top, bottom and middle point.
	%%
	We also extracted the variables for each projection axis to assess their information gain.
	}
  \label{fig:centerline}
\end{figure}
Four trained radiologists acquired the medical image data for each subject in a standardized way using a 1.5~Tesla Magnetic Resonance Imaging (MRI) scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
%%
The spine protocol consisted of a sagittal $T_1$- and $T_2$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
%%

@@

\paragraph{GPLOM Analysis.}

@@

\paragraph{Correlation Matrix.}

@@

\paragraph{Correlation of Image Variables With Lumbar Back Pain.}

@@

\paragraph{Assessing the Information Gain Using the Principal Component Analysis.}
%%
To determine the information gain per image-derived variable, we calculated a Principal Component Analysis (PCA) and compared the dimension loadings.
%%
The PCA results in a linear projection of the variable space ordering their components by maximum variance.
%%
The first three principal components explain 75\% of the variance in the nine image-derived variables.
%%
The first principal component explains 47\% of the variance and comprises primarily \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle}.
%%
The second component, adding 16\% of the variance, weigths \emph{coronal curvature} and \emph{coronal curvature angle}.
%%
The third component explains 12\% of the variance and comprises \emph{torsion} and \emph{transverse curvature}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%%
The complete analysis can be found in the experiments section of the supplementary material (\url{blind.dnsalias.com})
%% ------------------------------------------
\paragraph{Heterogenous Correlations.}
%%
We then expanded our focus on correlations of image-derived variables with all other non-image variables.
%%
Different correlation metrics depending on the type of the individual variables are used to derive correlations between all variables in the data set.
%%
The method uses the following correlation metrics for the different type combinations:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
%%
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - perfect correlation}.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
These variables have to be omitted in the analysis, since they are not statistically resilient.
%%
We display the resulting \emph{contingency matrix} using a heat map, encoding correlation values with color brightness with white for $0$ and dark blue for $1$, as presented in \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and searched for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the variable (see \url{blind.dnsalias.com}).
%%
Only weak correlations could be found for \emph{mean curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation of \emph{torsion} with \emph{Parkinson's disease} (0.24).
%%
Other than that, \emph{torsion} correlated with almost no variables (values between 0 and 0.05).
%%

@@

\subsection{The C4.5/C5.0 Algorithm}
\label{sec:C5.0}
The C4.5 algorithm builds decision trees based on information entropy on a data set.
%%
Such a calculation requires a numeric or categorical target variable.
%%
% The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
% \begin{algorithm}[]
%  Check for base cases\;
%  \For{each attribute $a$}{
%   Find the normalized information gain ratio from splitting on $a$\;
%  }
%  Let $a_{best}$ be the attribute with the highest normalized information gain\;
%  Create a decision node that splits on $a_{best}$\;
%  Iterate on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
%  \caption{Building a decision tree using the C4.5 Algorithm}
%  \label{alg:c4.5}
% \end{algorithm}
C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
%%
We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
%%
In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.

\subsection{Interactive Display of Decision Trees}
We have to create a decision tree for every non-image variable to analyze which one can be explained by image-derived variables.
%%
Since we have 134 non-image variables, the calculation yields the same amount of trees.
%%
Further subdivision, e.g. by quantiles of \emph{body size} increases the number to 402 trees.
%%
We have to abstract the results of the classification to keep the mental effort of interpreting the data low.
%%
This is included in our IVA framework.
%%------------
\subsubsection{Visualization of Classification Results}
\label{subsec:VisualizationOfClassificationResults}
%%
We follow the Visual Analytics mantra of analyzing first, show the important and analyze further \cite{Keim}.
%%
A first analysis step was performed by applying the classification algorithm to the data.
%%
The optimal classification uses a few rules to precisely characterize the target variable.
%%
Therefore, we are interested in \emph{small trees} with a \emph{low classification error}.
%%
The two measures form the axes for a scatter plot of the classification results.
%%
The scatter plot is our central element for the interactive analysis of decision trees.

\paragraph{The Error Term. }
Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
%%
The metric usually works well for variables with uniform distribution.
%%
It distorts the result for other distribution types.
%%
For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, the error metric would yield an error of 10\%, even though it is very bad.
%%
Based on this we use a summary error, which incorporates the discriminative power of each manifestation and is denoted as follows:
%%
\begin{equation}
totalError = 1 - \sum_{m=1}^M \frac{correctlyClassified_{m}}{M\cdot N_m}
\end{equation}
$M$ represents the set of manifestation of each variable.
%%
The error is scaled to denote perfect classification with 0; 1 is equal to random selection.
%%
We consider a result rated $0.5$ or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.
%%
It allows for comparability of error rates between variables with different manifestation count.

\paragraph{Attribute Mapping.}
%%
The scatter plot axes are defined by tree size and the previously described error metric.
%%
This allows us to visualize a multitude of classification results in one plot.
%%
Classification and comparison of variables for subject groups (e.g. male and female subjects) in one plot can be achieved by color coding group affiliation on the data points.

@@

The second plot shows all categorical variables transformed into dichotomous dummy variables (recall subsection~\ref{sec:C5.0}).

@@

Another observation during the analysis is that large decision trees achieve low error scores.
%%
Their complexity imposes overfitting to the data and they are not suitable for extracting universal rules.
%%
\paragraph{Applicability.}
%%
Classification methods based on decision trees have proven to be useful for assessing the discriminative power of a variable set.
%%
Their ability to consider variable combinations makes them more powerful than correlation coefficients calculated for each variable.
%%
This advantage comes with a much more complex output, the results are more challenging to assess and to abstract.
%%
Our method to plot derived metrics and custom-tailored error measures proved to be effective.
%%
Huge result spaces could be navigated fast using our interactive visualization.
%%
Therefore, the method is applicable not only for deriving information based on image data, but on all potential target variables.
%%

>>>>

\noindent Epidemiology is the study of dissemination, causes and results of health-related states and events.
%%
For this purpose, large population studies, such as the \emph{Study of Health in Pomerania} (SHIP) \cite{SHIP}, gather as much information as possible about participants to be assessed towards different diseases.
%%
This information is used to determine risk factors for diseases, help people to live healthier or to support the diagnosis of widespread diseases.
%%
% So far, epidemiological research is strongly hypothesis-driven.
% %%
% More precisely, observations made by clinicians are translated into hypotheses, which are then statistically evaluated using data variables from epidemiological studies.
% %%
% As a result, the hypothesis will be accepted or rejected.
% %%
%%
Cohort studies are an instrument of epidemiological research.
%%
Nowadays, these studies often comprise medical image data, e.g., tomographic image data sets. %, as well as general patient-specific attributes, such as \emph{gender}, \emph{age} and \emph{weight}.
%%
We analyze back pain, one of the most frequent complaints in the Western civilization. %\cite{Hoy2010}. % and gains importance due to the population aging \cite{Hoy2010}.
%%
Epidemiological analysis of back pain aims to determine risk factors.
%%
Although the shape and constitution of the spine, especially the lumbar spine, plays an important role for back pain, an automatic classification approach for characterization of back pain based on lumbar back spine attributes is still missing.
%%
%Furthermore, previous analyses mainly focused on a single variable (e.g. \emph{BMI}) or a small subset of variables.
%%

We present an analysis of the image-derived data from a cohort study lumbar spine dataset.
%%
We extract possible associations between spine shape and back pain characteristics.
%%
For this purpose, we combine data mining algorithms with data visualization techniques.
%%
Then, Interactive Visual Analysis (IVA) highlights mutual dependencies between image-derived data and back pain-related variables.
%%
% The workflow is part of an IVA framework, constructed to enhance the epidemiological workflow with iterative data-driven techniques.
%%
We focus on highlighting new correlations and trigger \emph{hypotheses generation}, rather than statistically validate complex epidemiological correlations.
%%
Our contributions are:
\begin{itemize}
\item An IVA workflow for back pain analysis based on image-derived variables of 2,240 subjects,
%%
\item The identification of lumbar spine shape properties potentially associated with back pain,
%%
\item The detection of associations between image-based, socio-demographic and medical variables for hypothesis generation,
%%
\item The identification of the most important variables via data mining methods using a novel semi- quantitative overview visualization concept. % embedded in our IVA-framework.
%%
\end{itemize}
%%
%The analysis, as well as the presented IVA tool is available as supplementary material in an interactive \texttt{R Markdown}\footnote{Developed by RStudio, Inc; \texttt{rmarkdown.rstudio.com}} document (\url{blind.dnsalias.com}).
%%
%% ---------------------------------------------
\section{\uppercase{Epidemiological Background}}
\label{sec:EpidemiologicalBackground}
%%
\noindent Epidemiological reasoning relies on a strict statistically-driven workflow \cite{Fletcher}:
%%
\begin{itemize}
%%
	\item Physicians formulate hypotheses based on observations made in their clinical practice.
%%
	\item Epidemiologists compile a list of variables depicting a hypothesis to prove it.
%%
	\item Statistical methods, such as regression analysis, assess the correlation of selected variables with the investigated disease.
%%
\end{itemize}
%%
Mutually dependent variables make this analysis challenging.
%%
Many diseases, such as different cancer types, are more likely with increasing age \cite{Fletcher}.
%%
% For example, the effect of nutrition on prostate cancer requires age normalization due to the increased incidence for older subjects.
% %%
% Age acts as a \emph{risk factor} for prostate cancer.
% %%
% Statistical correlation does not imply causation--epidemiologists need to assess the medical soundness of the statistical results.
%%------------------------------
\paragraph{Challenges of Epidemiological Data Analysis.}
% Epidemiological data originates from a wide range of studies.
% %%
% The study type depends on the condition of interest.
% %%
% Most common are case-control studies analyzing one specific disease and its influence on the human body.
%%
We focus on epidemiological large scale cohort study data.
%%
These studies aim at collecting as much data as possible for each subject.
%%
The collected data can be analyzed regarding many diseases and conditions.
%%
Epidemiological data are heterogenous and incomplete.
%%
For example, data about a disease treatment only affects subjects suffering from this condition.
%%
Therefore, statistical analysis has to take missing data into account.
%%
Epidemiological data acquisition includes a variety of techniques, such as medical examinations, self-reported questionnaires or genetic examinations.
%%
This yields a heterogenous information space.
%%
Information reduction techniques are necessary to compare these data.
%%
For example, continuous data, such as age, is often discretized into quantile bins for visualizing age-dependent disease frequencies.

Modern cohort studies often comprise medical image data.
The analysis of these data is challenging.
Individual diagnosis or manual segmentation of each body structure by radiologists is tedious, costly and comprises little reproducibility.
Segmentation algorithms are not generally available and need to be carefully adapted for each body structure.
%Segmentation data is usually analyzed by abstracting it into key figures, such as diameters or distances.
%These numeric values can be compared with non-image variables to retrieve correlations.
\\\\
\noindent \textbf{Lower Back Pain.}
Lower back pain is one of the most frequent complaints in the Western civilization \cite{Hoy2010}.
% Epidemiologists assume correlations between lumbar back pain and lifestyle factors.
% These include \emph{nutrition}, \emph{physical activity} and \emph{body posture combined with physical stress at work}.
The exact causes as well as particularly vulnerable risk groups are unknown.

@@

They have to analyze the lumbar spine shape as well as lifestyle factors.

@@

\noindent \textbf{Visual Analysis of Medical Image and Non-Image Data}

@@

In a previous work, we proposed an extension to the previously described epidemiological workflow using IVA \cite{Klemm2014VIS}.

@@

%We incorporate the IVA approach by concurrently displaying complex relationships between variables via decision trees.

@@

\\\\
\noindent \textbf{Visual Analysis of Heterogenous Non-Image Data.}

@@

% \cite{Dai} define a \emph{Concept Map}, linking cancer-associated features by incorporating choropleth maps with bar charts, parallel coordinates and scatter plots with regression lines.
% The \emph{Concept Map} is iteratively refined using primarily the underlying geographical data.

@@

\\\\
\noindent \textbf{Decision Rule-Driven Analysis of Medical Data.}

@@

\cite{Pinheiro} applied association rule mining to create high-confidence decision rules based on variables, such as age, gender and geographic location about the mortality rate in liver cancer data.
% \cite{Sekhavat} propose an interactive visualization technique for decision trees, where all decisions are displayed in a heat map.
% The visualization translates the decision tree into a heat map, where each cell represents the connectivity between an initial and a target value.
% The resulting interactive visualization is well suited for analyzing single, but not multiple decision trees.
\\\\
\noindent \textbf{SHIP Data Analysis.}
\cite{Klemm2013VMV} analyzed the lumbar spine variability of 490 \texttt{SHIP-2} subjects.
They incorporate hierarchical agglomerative clustering to derive shape groups, yielding several groups of average shape and several outlier clusters.
The extracted lumbar spine shape was correlated with subject size.
Our data extraction process is based on the detection algorithm and lumbar spine canal extraction presented by \cite{Klemm2013VMV}.
Clustering techniques are strongly dependent on the chosen variables and distance measures \cite{Klemm2014BVM}.
\\\\
Unique in our work is the combination of data mining techniques with an IVA approach by observing interesting feature relations in the context of image-derived features using multiple decision trees for one visualization.
We abstract the decision tree results similar to \cite{Turkay}, making them comparable in an overview visualization.
\section{\uppercase{The Lumbar Spine Data Set}}
\label{sec:MaterialsAndMethod}
\noindent Our approach allows the simultaneous analysis of many variables.
Therefore, epidemiologists compiled the data set with a wide range of variables possibly correlating with lumbar back pain.
The data set comprises 6,753 subjects from two cohorts (4,420 from \texttt{SHIP-Trend-0} and 2,333 from \texttt{SHIP-2}) including non-image and image-derived data.
\\\\
\noindent \textbf{Non-Image Data}.
The non-image variables range from somatometric variables describing body measures to medical examinations, such as laboratory tests as well as lifestyle factors, e.g. sporting activity or nutrition.
The data set comprises 134 variables (and 9 additional image-derived variables described in the next subsection):
	\item 21 metric variables, mostly describing somatometric variables and markers retrieved from blood analysis
	\item 113 categorical variables divided into
		\item 43 dichotomous (binary) variables, mostly indicating the presence of a disease, e.g. \emph{pancreatitis} or \emph{high blood pressure}
		\item 70 variables with more than two levels, indicating pain levels, nutrition and social factors, such as marital or retirement status
Many variables comprise missing data.
Some variables are follow-up questions, covering the treatment of a disease.
Others are exclusive to subgroups, such as women-specific variables, e.g. \emph{menstrual status}.
Sparse variables are statistically less resilient and need to be treated with care during the analysis.
\\\\
\noindent \textbf{Image-Derived Data.}
\begin{figure}[!t]
  \includegraphics[width=0.475\textwidth]{figures/centerline}
	(a) left: Tetrahedron-based finite element model (FEM) of the lumbar spine, capturing the L1-L5 vertebrae as well as the lumbar spine canal.
	(a) right: FEM aligned to an MRI data set.
	The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
	(b) Using the Frenet frame, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).
	The curvature angle ($\alpha$) is calculated using the top, bottom and middle point.
	We also extracted the variables for each projection axis to assess their information gain.
  \label{fig:centerline}
\end{figure}
Four trained radiologists acquired the medical image data for each subject in a standardized way using a 1.5~Tesla Magnetic Resonance Imaging (MRI) scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
The spine protocol consisted of a sagittal $T_1$- and $T_2$-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.

@@

\\\\
\noindent \textbf{GPLOM Analysis.}

@@

\\\\
\noindent \textbf{Correlation Matrix.}

@@

\\\\
\noindent \textbf{Correlation of Image Variables With Lumbar Back Pain.}

@@

\\\\
\noindent \textbf{Assessing the Information Gain Using the Principal Component Analysis.}
To determine the information gain per image-derived variable, we calculated a Principal Component Analysis (PCA) and compared the dimension loadings.
The PCA results in a linear projection of the variable space ordering their components by maximum variance.
The first three principal components explain 75\% of the variance in the nine image-derived variables.
The first principal component explains 47\% of the variance and comprises primarily \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle}.
The second component, adding 16\% of the variance, weigths \emph{coronal curvature} and \emph{coronal curvature angle}.
The third component explains 12\% of the variance and comprises \emph{torsion} and \emph{transverse curvature}.
This supports our prior conclusion about the low information gain of the transverse planes.
\emph{Torsion} also adds little variance to the information space.
The complete analysis can be found in the experiments section of the supplementary material (\url{blind.dnsalias.com})
%% ------------------------------------------
\\\\
\noindent \textbf{Heterogenous Correlations.}
We then expanded our focus on correlations of image-derived variables with all other non-image variables.
Different correlation metrics depending on the type of the individual variables are used to derive correlations between all variables in the data set.
The method uses the following correlation metrics for the different type combinations:
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - perfect correlation}.
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
These variables have to be omitted in the analysis, since they are not statistically resilient.
We display the resulting \emph{contingency matrix} using a heat map, encoding correlation values with color brightness with white for $0$ and dark blue for $1$, as presented in \cite{Klemm2014VIS}.
We calculated the contingency matrix for all size groups and searched for correlations between image- and non-image derived variables.
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the variable (see \url{blind.dnsalias.com}).
Only weak correlations could be found for \emph{mean curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
One surprising result was the small correlation of \emph{torsion} with \emph{Parkinson's disease} (0.24).
Other than that, \emph{torsion} correlated with almost no variables (values between 0 and 0.05).

@@

We use the C4.5 algorithm, which builds decision trees based on information entropy on a data set.
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
%%
In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.

%\subsection{The C4.5/C5.0 Algorithm}
%\label{sec:C5.0}
%The C4.5 algorithm builds decision trees based on information entropy on a data set.
%%
%Such a calculation requires a numeric or categorical target variable.
%%
% The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
% \begin{algorithm}[]
%  Check for base cases\;
%  \For{each attribute $a$}{
%   Find the normalized information gain ratio from splitting on $a$\;
%  }
%  Let $a_{best}$ be the attribute with the highest normalized information gain\;
%  Create a decision node that splits on $a_{best}$\;
%  Iterate on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
%  \caption{Building a decision tree using the C4.5 Algorithm}
%  \label{alg:c4.5}
% \end{algorithm}
%C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
%%
%We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
%%
% Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
% %%
% Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
% %%
% In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.

\subsection{Interactive Display of Decision Trees}
We have to create a decision tree for every non-image variable to analyze which one can be explained by image-derived variables.
%%
Since we have 134 non-image variables, the calculation yields the same amount of trees.
%%
Further subdivision, e.g. by quantiles of \emph{body size} increases the number to 402 trees.
%%
We have to abstract the results of the classification to keep the mental effort of interpreting the data low.
%%
This is included in our IVA framework.
%%------------
\subsubsection{Visualization of Classification Results}
\label{subsec:VisualizationOfClassificationResults}
%%
We follow the Visual Analytics mantra of analyzing first, show the important and analyze further \cite{Keim}.
%%
A first analysis step was performed by applying the classification algorithm to the data.
%%
The optimal classification uses a few rules to precisely characterize the target variable.
%%
Therefore, we are interested in \emph{small trees} with a \emph{low classification error}.
%%
The two measures form the axes for a scatter plot of the classification results.
%%
The scatter plot is our central element for the interactive analysis of decision trees.
\\\\
\noindent \textbf{The Error Term. }
Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
%%
The metric usually works well for variables with uniform distribution.
%%
It distorts the result for other distribution types.
%%
For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, the error metric would yield an error of 10\%, even though it is very bad.
%%
Based on this we use a summary error, which incorporates the discriminative power of each manifestation and is denoted as follows:
%%
\begin{equation}
totalError = 1 - \sum_{m=1}^M \frac{correctlyClassified_{m}}{M\cdot N_m}
\end{equation}
$M$ represents the set of manifestation of each variable.
%%
The error is scaled to denote perfect classification with 0; 1 is equal to random selection.
%%
We consider a result rated $0.5$ or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.
%%
It allows for comparability of error rates between variables with different manifestation count.
\\\\
\noindent \textbf{Attribute Mapping.}
%%
The scatter plot axes are defined by tree size and the previously described error metric.
%%
This allows us to visualize a multitude of classification results in one plot.
%%
Classification and comparison of variables for subject groups (e.g. male and female subjects) in one plot can be achieved by color coding group affiliation on the data points.

@@

The second plot shows all categorical variables transformed into dichotomous dummy variables. % (recall subsection~\ref{sec:C5.0}).

@@

Another observation during the analysis is that large decision trees achieve low error scores.
%%
Their complexity imposes overfitting to the data and they are not suitable for extracting universal rules.
%%
\\\\
\noindent \textbf{Applicability.}
%%
Classification methods based on decision trees have proven to be useful for assessing the discriminative power of a variable set.
%%
Their ability to consider variable combinations makes them more powerful than correlation coefficients calculated for each variable.
%%
This advantage comes with a much more complex output, the results are more challenging to assess and to abstract.
%%
Our method to plot derived metrics and custom-tailored error measures proved to be effective.
%%
Huge result spaces could be navigated fast using our interactive visualization.
%%
Therefore, the method is applicable not only for deriving information based on image data, but on all potential target variables.
%%

% BLIND
%As described in a prior work (\cite{Klemm2014VIS}), the suggestion of potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
Suggesting potentially interesting features when analyzing a condition is an important aspect of visual analytics in epidemiology.
%%
We can achieve this also using the presented method by trying to describe several (dichotomous) target values with a decision tree constructed from all available data.
%%
This allows to assess both the discriminative power of the data set regarding the target variables, and as the most important variables for the classification.
%%
\\\\
\noindent \textbf{Predictive Power of Image-Derived Variables.}
We showed that \emph{torsion}, \emph{curvature} and \emph{curvature angle} of the lumbar spine at the presented precision are not sufficient to characterize lumbar back pain in the SHIP data set.
%%
Our method allowed to assess their discriminative power, which is largely limited for separating male and female subjects, \emph{nutrition} variables, as well as different disease indicators.
%%
The C4.5 algorithm proved to be an effective tool for evaluating a set of derived metrics regarding their suitability to classify non-image variables.
%%
Over-fitting to the data indicated by complex decision trees has to be taken into account as well.
%%
\\\\
\noindent \textbf{Future Work.}
%% Contributions
We provide a method for comparative analysis of decision trees independent of the variable manifestation count using interactive scatter plots.
%%
We applied the technique to gain insight into the predictive power of 9 image-derived variables for 134 non-image variables.
%%
The analysis was carried out for subjects divided by \emph{gender}, \emph{BMI}, and \emph{body size} to assess their influence on the lumbar spine shape.
